Solution of an Acoustic Transmission Inverse Problem by Extended Inversion

### 2020-08-19 repro/papers/discrep/paper.tex
interim

\title{A Discrepancy-Based Algorithm for Extended Inversion}
\author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}

@@

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  recorded at a given distance from a point source. The velocity or
  slowness is presumed homogeneous, and the target source wavelet is
  presumed quasi-impulsive or localized near zero time lag. The source
  is extended by permitting energy to spread in time, and the spread
  is controlled by adding a weighted mean square of the extended
  source wavelet to the data misfit, to produce the extended inversion
  objective. The objective function and its gradient can be computed
  explicitly, and it is easily seen that the error in any local
  minimizer is bounded by a multiple of the permitted energy spread,
  which may be interpreted as a wavelength. The derivation shows several
  important features of all similar extended source algorithms. For
  example, nested optimization, with the source estimation in the
  inner optimization (variable projection method), is essential. The
  choice of the penalty operator, controlling the extended source
  degrees of freedom, is critical: in order to produce an objective
  immune from cycle-skipping, the penalty operator must be
  (pseudo-)differential. The penalty weight is chosen dynamically,
  according to a version of the Discrepancy Principle, in order to
  accelerate convergence to a near-global minimum. Like many other
  extended formulations, the example problem considered here permits
  the discrepancy criterion to be satisfied with zero penalty
  weight. This unusual feature removes a common obstacle to the use of
  the discrepancy principle, namely the determination of a
  satisfactory estimate of the penalty weight: the initial weight may
  be set to zero, and the dynamic algorithm updates it until
  convergence is achieved.

@@

\section{Introduction}
Full Waveform Inversion (FWI), or estimation of earth structure by
model-driven least squares data fitting, is now well-established as a
useful tool for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10}. However, so-called ``cycle-skipping'', the tendency of iterative FWI
algorithms to stagnate at suboptimal and geologically uninformative
earth models, still impedes its use. Because the computational size of field inversion tasks
is very large, only iterative local (descent) minimization of the data
misfit function is computationally feasible. However local
descent methods avoid suboptimal stagnation only if initial models are
already quite close to optimal, in the sense of predicting the arrival
times of seismic events to within a small multiple of a dominant
wavelength \cite[]{GauTarVir:86,Plessix:10}.

This paper concerns one of the several ideas that have been advanced to
overcome cycle-skipping, namely so-called extended inversion
\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
of freedom are provided to the modeling process, in the hope of
opening up more effective routes to geologically informative models
with acceptable data fit. Since these extended degrees of freedom are
not part of the basic physics chosen to model the data acquisition
process, they should be suppressed in the eventual solution. Extended
inversion methods differ by the choice of additional degrees of
freedom, and by choice of penalty applied to eliminate them in the final
result.

Many of these extended inversion concepts sound plausible, and appear
to work at least to some extent as one might hope from their heuristic
justifications. However very few of these approaches have been
underwritten by mathematical argument: in essence, they are mostly
justified only ``in the rear-view mirror'', with no assurance that
failure is not just around the corner, at the next example. On top of
that, some of these approaches, for example those based on the
computationally attractive Variable Projection Method (``VPM'') of
\cite{GolubPereyra:03}, are cast in such form that the reasons for
success are not readily apparent.

This note shows exactly how extended inversion leads to successful solution
of a very simple wave propagation inverse
problem, which asks that a homogeneous velocity field be deduced from
one trace at known offset. I put forward this inverse problem and
extension-based solution not because there are not simpler
ways of answering the question it poses - there certainly are - but
because the formal ingredients of waveform-based velocity estimation
in this very simple setting are common to many similar extended
inversion algorithms, and because in this case every computation can
be done analytically, nearly to completion. In particular, it becomes
clear why the VPM gradient formula produces a constructive update,
with no possibility of stagnation away from the global minimum.

The extended inversion approach
developed here uses a {\em source extension}, in which source
parameters form the additional degrees of freedom. This type of
extension presumes that the actual or target source is constrained in
some way; the extended source is allowed to violate the
constraint. For recent overview of source extension methods, see
\cite{HuangNammourSymesDollizal:SEG19}. Source
extension methods have computational complexity approximately the same
as that of FWI, a signal advantage over the alterantive medium
extension class described for instance in \cite[]{geoprosp:2008}.

For the problem considered here, the source model amounts to a
wavelet, and the target wavelet is assumed to be non-zero only in a
short time interval (an approximate impulse, perhaps as the result of
signature deconvolution). The extension consists in permitting energy
to spread in time at intermediate iterations of the inversion. A
penalty for energy spread drives the extended source towards a focused
source approximately satisfying the assumed constraint. Not all
penalties are created equal. Two plausible penalty operators are
studied here, both multiplication operators: one by the characteristic
function of the target support, the other by a smooth function of
time. Only the smooth multiplier avoids cycle-skip. This distinction
is quite general, as will be explained in the Discussion section.

%despite not incorporating explicitly the extension measure
%(``annihilator'') central to the approach.

I begin with a quick sketch of constant density acoustics, and
describe the single-trace transmission inverse problem.. For
completeness, I show how the standard FWI approach to this problem,
with the small support of the source imposed as a hard constraint,
generates multiple local minima that will be found by any descent
method unless the initial slowness estimate predicts travel time from
source to receiver with an error on the order of a wavelength. The
next section describes the extended source objective, and the reduced
objective produced by VPM, in general terms. As VPM eliminates the
extended source, this function depends only on the velocity, just as
does the FWI objective.  Of course, it also depends on the operator
used to define the penalty (``annihilator''). I propose two choices: a
soft version of the support constraint, and an even softer constraint,
namely multiplication by time. A nearly-explicit calculation of the
VPM gradient for these two choices shows that the first leads again to
multiple local minima far from the target solution, whereas the second
does not. In fact, in the second case, the only stationary points are
``within a wavelength'' of the correct velocity, used to build the
data: that is, cycle-skipping cannot occur. For this second, more
successful penalty, I introduce the discrepancy based dynamic
weighting algorithm, and show how it accelerates the convergence
towards a local minimizer, which has been shown to be an acceptable
approximation to a global minimizer. The paper ends with a discussion
of the parallels between the calculations presented here and the
structure of other extended inversion methods applicable to
field-scale velocity estimation, and the critical role that the
differential nature of the extension penalty plays in the success of
this and other extension methods.

I shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for
the algebra of bounded linear operators from the Hilbert space $X$ to
the Hilbert space $Y$, and its subalgebra of invertible operators.

%\section{Theory}

\section{An Example: Single Trace Modeling}
Assume small amplitude (linearized) acoustic propagation, constant
density, and isotropic point source and receiver. Denote by $m(\bx)$
the slowness (reciprocal velocity) at spatial position $\bx$, $w(t)$
the time dependence of the point source (``wavelet'') at location
$\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
scalar wave equation:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m(\bx)^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0
\end{eqnarray}
Suppose that a single trace is recorded, at distance $r>0$ from the
source position $\bx_s$. The dominant information in a single
trace is the transient signal time of arrival, constraining only the mean slowness in the
region
between source and receiver, so assume that the
slowness is constant, that is, independent of position $\bx$. The pressure field is simply the  the source
wavelet $w(t)$ convolved with the
acoustic Green's function, for which an analytic expression is
available in the constant $m$ case \cite[]{CourHil:62}:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right).
\end{equation}

The receiver location $\bx_r$ lies at distance $r$ from the source
location $\bx_s$, that is, $|\bx_r-\bx_s|=r$. The predicted signal at
$p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
source wavelet $w$. Therefore it is naturally represented as the
action of a $m$-dependent linear operator $S[m]$ on $w$:
\begin{equation}
\label{eqn:mod}
F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
\end{equation}

Ignoring amplitude, this map implements a $m$-dependent time
shift. This time shift operator is the basis of many descriptions of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above, which
amounts essentially to a time shift.

The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range $M=(m_{\rm min}, m_{\rm max})$ characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km. The wavelet $w$ is
naturally assumed to be 
square-integrable, corresponding to finite power transfer from the
source process to the propagating wavefield $p$
\cite[]{SantosaSymes:00}. The predicted data $F[m]w$ is then also
square-integrable.

FIeld recording takes place over finite time intervals (and with
finite sample rate, a detail I shall ignore in this paper). Denote by
$[t_{\rm min},t_{\rm max}]$ the recording interval for the
hypothetical single trace data. Then natural choices for domain and
range of $F[m]$ are $W = L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm
  max}])$ respectively (that is, interpret the definition \ref{eqn:mod}
as holding for $t \in [t_{\rm min},t_{\rm  max}]$). So
\begin{equation}
  \label{eqn:mapprop}
  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
  \frac{1}{4\pi r}.
\end{equation}

\section{The Inverse Problem}

@@

Note that for $w \in \lW$, $m \in M$,
$\mbox{supp }F[m]w \subset [ m_{\rm min}r-\lambda,
m_{\rm max}r+\lambda] \cap
[t_{\rm min},t_{\rm max}]$. Also, $\lF[m]$ is coercive for all $m \in
M$ (so that $w$ is uniquely determined) only if
$[ m_{\rm min}r-\lambda,m_{\rm max}r+\lambda] \subset
[t_{\rm min},t_{\rm max}]$. Define $\lambda_{\rm max}
>0$, so that $M, r,$ and $T$ by
\begin{equation}
  \label{eqn:fullrec}
  [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
  \subset [t_{\rm min},t_{\rm max}].
\end{equation}
Then for 
$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $F[m] \in {\cal B}(\lW,D)$ is coercive.

@@

\noindent {\bf Problem Statement:} given data $d \in D$, relative 
error level $\epsilon >0$, and support radius $0 <\lambda \le 
\lambda_{\rm max}$, find 
$(m,w) \in M \times \lW$ for which 
\begin{equation}
  \label{eqn:probstat0}
  \|\lF[m]w-d\| \le \epsilon\|d\|. 
\end{equation}

\noindent{\bf Remark:} The support constraint is closely linked to the
folk theorem about FWI noted many times in the literature: convergence
of a descent method requires that the initial slowness must be known
to ``within a (fraction of a) wavelength''. The relation is a
consequence of Heisenberg's inequality, and will be reviewed later in
the paper.

The best case for data fitting
is clearly the one in which the data can be fit precisely: that is,
there exists $(m_*,w_*) \in M\times \lW$ so that
\begin{equation}
  \label{eqn:defdatanonoise}
  d=\lF[m_*]w_*.
\end{equation}
Such data $d$ is {\em noise-free}, in the range of the map $\lF$, and a solution of the Problem Statement
exists with arbitrarily small $\epsilon>0$. I will relax the noise-free assumption
later in the paper: until further notice, assume that $d \in
D$ satisfies the condition \ref{eqn:defdatanonoise}.

\section{Full Waveform Inversion: hard constraint}
A natural formulation of the Problem Statement \ref{eqn:probstat0} is
via least squares: that is, given $d \in D$, $\lambda >0$, find $(m,w)
\in M \times \lW$ to minimize $e: M \times W
 \rightarrow \bR$, defined by
\begin{equation}
  \label{eqn:FWI}
  e[m,w;d] = \frac{1}{2}\|\lF[m]w-d\|^2.
\end{equation}
A pair $(m,w) \in M\times \lW$
satisfies the condition \ref{eqn:probstat0} iff $e[m,w;d] \le \epsilon^2/2$, so a solution exists iff a 
global minimizer of $e$ satisfies this inequality.

>>>>

\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
\author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}

@@

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  recorded at a given distance from a point source. The velocity or
  slowness is presumed homogeneous, and the target source wavelet is
  presumed quasi-impulsive or localized near zero time lag. The source
  is extended by permitting energy to spread in time, and the spread
  is controlled by adding a weighted mean square of the extended
  source wavelet to the data misfit, to produce the extended inversion
  objective. The objective function and its gradient can be computed
  explicitly, and it is easily seen that the error in any local
  minimizer is bounded by a multiple of the permitted energy spread,
  which may be interpreted as a wavelength. The derivation shows several
  important features of all similar extended source algorithms. For
  example, nested optimization, with the source estimation in the
  inner optimization (variable projection method), is essential. The
  choice of the penalty operator, controlling the extended source
  degrees of freedom, is critical: in order to produce an objective
  immune from cycle-skipping, the penalty operator must be
  (pseudo-)differential. The penalty weight is chosen dynamically,
  according to a version of the Discrepancy Principle, in order to
  accelerate convergence to a near-global minimum. Like many other
  extended formulations, the example problem considered here permits
  the discrepancy criterion to be satisfied with zero penalty
  weight. This unusual feature removes a common obstacle to the use of
  the discrepancy principle, namely the determination of a
  satisfactory initial estimate of the penalty weight: the initial weight may
  be set to zero, and the dynamic algorithm updates the weight until
  convergence is achieved.

@@

\section{Introduction}
Model-based parameter estimation by least squares data fitting is a
widely used and successful technique for data analysis in science and
engineering \cite[]{Bertero:98,Vogel:02}, particularly in the
geosciences \cite[]{Parker:94,Tarantola:05}. In its application to
seismology, least-squares data fitting has come to be known as Full
Waveform Inversion (FWI), and is now well-established as a useful tool
for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
application to seismology, FWI encounters a serious
practical obstacle, so-called ``cycle-skipping''. This term signifies
the tendency of the mean-square error function to exhibit many
stationary points, and local minima far from the global minimizer (we
show an explicit example of this phenomenon later in this paper). Because of the
typical dimensions of field FWI problems, only local gradient-based
minimization algorithms are computationally practical, and these find
local minimizers, so may stagnate at suboptimal and geologically
uninformative earth models. Local descent methods avoid suboptimal
stagnation only if initial models are already quite close to optimal,
in the sense of predicting the arrival times of seismic events to
within a small multiple of the dominant data wavelength
\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.

This paper examines one of a number of proposed alternatives to least squares
FWI, based on an extension of the model-to-data mapping.  We show
exactly how an example of this {\em extended inversion} approach leads
to successful solution of a very simple acoustic transmission inverse
problem. Despite its simplicity, this problem exhibits several
important characteristics of all inverse problems in wave propagation
in which wave velocities are amongst the unknowns. It has the
advantage of permitting explicit calculations that reveal clearly the
failure mechanism for FWI, that is, straightforward nonlinear least
squares, and also the reasons for success of an extended inversion
approach.

In the remainder of this introduction, we will briefly describe the conceptual
framework, the inverse problem, and sketch our main results. We leave out
some technical details that are supplied in the next section,
containing precise statements of the results.

The idealized physical setting is the propagation of wave through a
homogeneous acoustic fluid occuyping $\bR^3$ \cite[]{Frie:58}. The
fluid's salient characteristic for the present discussion is the
slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
energy localized at a point in $\bR^3$ radiates uniformly in all
directions with time-dependent intensity $w(t)$. Both the acoustic
pressure field and the intensity (``wavelet'') $w$ vanish at large
negative time. A receiver records the pressure values at a distance
$r>0$ from the source position, between the times $t_{\rm min}$ and
$t_{\rm max}$. As explained in the next section, the
recorded presssure is predicted to be
\begin{equation}
\label{eqn:mod}
\frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
\end{equation}
Ignoring the amplitude factor $1/(4\pi r)$, this map implements a $m$-dependent time
shift. This time shift operator is the basis of many descriptions of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above.

Evidently if the wavelet $w$ is square-integrable, so is the predicted
recorded pressure - its $L^2$ norm is the cumulative power trasferred
from the fluid to the receiver \cite[]{SantosaSymes:00}. Thus this relation defines a map
$F: \bR^+ \times L^2(\bR) \rightarrow
L^2([t_{\rm min},t_{\rm max}])$, linear in its second argument.

A naive version of the inverse problem investigated in this paper
would be: given recorded data
$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR$ and $w \in L^2(\bR)$ so that
$F[m]w \approx d$. Clearly this problem statement is not interesting as it
stands, since it is always possible to solve it with no error and nonuniquely,
and it does not constrain the slowness $m$ at all. To be useful, the statement must
be augmented with a constraint on the model $(m,w)$. We choose 
$\lambda > 0$ and search for $w$ with $\mbox{supp }w \subset
[-\lambda,\lambda]$. The quantity $\lambda$ is to be considered
``small'': as we shall show, it is related to a mean wavelength of
$w$.

Abusing notation slightly, we identify $L^2([-\lambda,\lambda])$ with
a subspace of $L^2(\bR)$ via extension by $0$. Denote by $\lF$
restriction of $F$ to $\bR^+ \times L^2([-\lambda,\lambda])$ defined by equation
\ref{eqn:mod}. Then the FWI problem with support constraint,
described in the last paragraph, can be crudely phrased: given $d \in
L^2([t_{\rm min},t_{\rm max}])$, find $(m,w)$ in the domain of $\lF$ so that $\lF[m,w]
\approx d$.

Despite its simplicity, this problem exhibits the
cycle-skipping pathology characteristic of field-scale FWI. Specifically, we show that the reduced mean square
error function, the objective function of FWI for this problem, 
\begin{equation}
  \label{eqn:redms}
  (m,w) \mapsto \lerr[m,w,d]=\frac{1}{2}\|\lF[m]w-d\|^2,
\end{equation}
has local minimizers far from any global minimizer, even for noise-free
data (Theorem \ref{thm:fwi}).

This fact implies that local optimization must fail to solve a precise
version of the FWI problem: given $\epsilon, \lambda > 0$, $(m,w)$ is
a solution if it lies in the domain of $\lF$ and
$\lerr[m,w,d] \le 1/2 \epsilon^2\|d\|^2$ (Problem Statement
\ref{eqn:probstat0} below).  In fact, we show
that``large residual'' local minima of $\lerr[\cdot,\cdot,d]$ exist,
at which its value is much greater than its global minimum
value. Therefore if $\epsilon$ is near the global minimum value, then
local minimization of $\lerr[\cdot,\cdot,d]$ can fail to find a
solution of the inverse problem. This phenomenon is what is meant by
``cycle-skipping''.

Extended inversion involves an extension of the model-to-data map, in
this case $\lF$.  The extension used here consists in dropping the
support constraint on $w$, so the extended map is simply $F$ as
defined above ($\lF$ is a restriction of $F$, so $F$ is an extension
of $\lF$). To compensate for the resulting indeterminacy in the
estimation of $m$, we add a penalty term to the objective: for $m \in
\bR^+,w\in L^2(\bR), d \in L^2([t_{\rm min},t_{\rm max}])$, define
\begin{eqnarray}
  \label{eqn:pen}
  e[m,w;d] & = & \frac{1}{2}\|F[m]w-d\|^2,\\
  g[w] & = & \frac{1}{2}\|A\|^2, \\
  \Ja[m, w;d] &=& e[m,w;d] + \alpha^2 g[w] \nonumber\\
  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2). 
\label{eqn:pen}
\end{eqnarray}
The only obvious requirement on the penalty operator $A$ is that it be
nonzero for large $|t|$, so that minimizers of $\Ja$ should be
concentrated near $t=0$. Thus the penalty is a soft stand-in for the
support constraint. We
show that a good choice for $A$ is given by $(Aw)(t)=tw(t)$, and the
remainder of the discussion in this section makes this assumption. The
penalty weight $\alpha$ must be chosen somehow; we review a simple
algorithm for controlling $\alpha$ later in the paper.

Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*w_*]$
the noise-to-signal ratio. We show that any local minimizer $(m,w)$
of $\Ja$ satisfies 
\begin{itemize}
\item If $\eta \le 0.6$,
  \begin{equation}
    \label{eqn:mbnd}
    |m-m_*| \le (1+f(\eta))\frac{\lambda}{r},
  \end{equation}
  where $f(\eta)=2\eta+O(\eta^2)$ (Theorem \ref{thm:mnoiseres});
\item 
  $(m,{\bf 1}_{[-\lambda,\lambda]}w)$ is a solution of the FWI problem
  as stated above, that is,
  \begin{equation}
    \label{eqn:success}
    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w,d] \le \frac{1}{2}\epsilon^2,
  \end{equation}
  so long as
  \begin{eqnarray}
    \lambda & \ge & (2+f(\eta))\mu,\nonumber\\
    \epsilon &\ge & (4\pi r \alpha \lambda)^2 +\eta.
    \label{eqn:solcond}
  \end{eqnarray}
  (Theorem \ref{thm:ipnoisesuf})
\end{itemize}.
That is, the minimization of the extended objective $\Ja$ provides a
solution of the FWI problem, with constraints on the achievable
performance and bounds on the errors. All of these
bounds are shown to be sharp, and the constraints cannot be
significantly weakened. For example, the support radius $\lambda$ in
the projection onto $L^2([-\lambda,\lambda])$ must in general be at least twice the
support radius of the ``noise free'' wavelet $w_*$, plus an additional
allowance for noise.

The remainder of the paper begins with a complete description of the acoustic transmission problem and precise
statements of our main results, filling in the
few details not discussed in the preceding sketch. Then we provide an
discussion of the potential theoretical and practical import of these
results, including relations with prior work, some numerical
illustrations, and directions for further research. Following this
discussion section, we give complete proofs of all of our results.
An appendix describes the computational algorithm used in producing
the numerical examples, including a method for controlling the penalty
weight $\alpha$ based on the Discrepancy Principle
\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}. A second appendix
explains the relation between the support constraint and well-known
criteria for success of FWI based on data frequency content. The basis
for this relation is the Heisenberg inequality \cite[]{Folland:07}.

\section{An Acoustic Transmission Inverse Problem}
In this section, we describe solution of the radiation
problem for acoustic wave
propagation in a homogeneous fluid, and various operators based on
this solution. Using these operators, we formulate the single-trace
acoustic transmission inverse problem. Finally, we state the main
results of our paper.

We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
the algebra of bounded linear operators from the Hilbert space $X$ to 
the Hilbert space $Y$, and its subalgebra of invertible operators.

\subsection{Acoustic transmission}
Assume small amplitude (linearized) acoustic propagation, constant
density, and isotropic point source and receiver. Denote by $m(\bx)$
the slowness (reciprocal velocity) at spatial position $\bx$, $w(t)$
the time dependence of the point source (``wavelet'') at location
$\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
scalar wave equation:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m(\bx)^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0
\end{eqnarray}
Suppose that a single trace is recorded, at distance $r>0$ from the
source position $\bx_s$. The dominant information in a single
trace is the transient signal time of arrival, constraining only the mean slowness in the
region
between source and receiver, so assume that the
slowness is constant, that is, independent of position $\bx$. The pressure field is simply the  the source
wavelet $w(t)$ convolved with the
acoustic Green's function, for which an analytic expression is
available in the constant $m$ case \cite[]{CourHil:62}:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right).
\end{equation}

The receiver location $\bx_r$ lies at distance $r$ from the source
location $\bx_s$, that is, $|\bx_r-\bx_s|=r$. The predicted signal at
$p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
source wavelet $w$. Therefore it is naturally represented as the
action of a $m$-dependent linear operator $F[m]$ on $w$, as described
in \ref{eqn:mod}.
%\begin{equation}
%\label{eqn:mod}
%F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
%\end{equation}

The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range $M=(m_{\rm min}, m_{\rm max})$ characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.

Field recording takes place over finite time intervals (and with
finite sample rate, a detail we shall ignore in this paper). Denote by
$[t_{\rm min},t_{\rm max}]$ the recording interval for the
hypothetical single trace data. Then natural choices for domain and
range of $F[m]$ are $W = L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm
  max}])$ respectively (that is, interpret the definition \ref{eqn:mod}
as holding for $t \in [t_{\rm min},t_{\rm  max}]$). So
\begin{equation}
  \label{eqn:mapprop}
  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
  \frac{1}{4\pi r}.
\end{equation}
Note also that $F[m]$ is surjective for every $m \in M$.

In computational practice, $W$ will have to be replaced by a
finite-dimensional subspace of $L^2(\bR)$. Many such choices will
implicitly limit the support of $w \in W$ to a bounded interval, say
$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
\begin{eqnarray}
  \label{eqn:dombounds}
  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
\end{eqnarray}
We will ignore these computational necessities in this work,
maintaining the definition $W=L^2(\bR)$.

\subsection{An inverse problem}

@@

Note that for $w \in \lW$, $m \in M$,
$\mbox{supp }F[m]w \subset [ m_{\rm min}r-\lambda,
m_{\rm max}r+\lambda] \cap
[t_{\rm min},t_{\rm max}]$. Also, $\lF[m]$ is coercive for all $m \in
M$ only if
$[ m_{\rm min}r-\lambda,m_{\rm max}r+\lambda] \subset
[t_{\rm min},t_{\rm max}]$.

We will assume that $\lambda_{\rm max}
>0$ is chosen so that $M, r,\lambda_{\rm max}$ and $[t_{\rm min},t_{\rm max}]$ satisfy
\begin{equation}
  \label{eqn:fullrec}
  [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
  \subset [t_{\rm min},t_{\rm max}].
\end{equation}
Then for 
$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $\lF[m] \in {\cal B}(\lW,D)$ is coercive.

@@

\noindent {\bf Problem Statement:} given data $d \in D$, relative 
error level $\epsilon \in [0,1)$, and support radius $\lambda \in (0, 
\lambda_{\rm max}]$, find 
$(m,w) \in M \times \lW$ for which 
\begin{equation}
  \label{eqn:probstat0}
  \|\lF[m]w-d\| \le \epsilon\|d\|. 
\end{equation}

\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
target noise level is eliminates the obvious choice $(m,0)$, which
satisfies the data misfit constraint for any $m \in M$.

\noindent{\bf Remark:} The support constraint is closely linked to the
folk theorem about FWI noted many times in the literature: convergence
of a descent method requires that the initial slowness must be known
to ``within a (fraction of a) wavelength''. The relation is a
consequence of Heisenberg's inequality, and will be reviewed in
Appendix B.

The best case for data fitting
is clearly the one in which the data can be fit precisely: that is,
there exists $(m_*,w_*) \in M\times \lW$ so that
\begin{equation}
  \label{eqn:defdatanonoise}
  d=\lF[m_*]w_*.
\end{equation}
Such data $d$ is {\em noise-free}, in the range of the map $\lF$. For
such data a solution of the Problem Statement
exists with arbitrarily small $\epsilon>0$.

\subsection{Soft constraints and penalty formulation}

As mentioned in the introduction, a straightforward nonlinear least
squares formulation of the problem stated in \ref{eqn:probstat0}
exhibits local minimizers with large fit error, and these will be
found by any local optimization algorithm unless it is initialized at
a quite accurate initial model. The first result stated in the next
section establishes this assertion for the single trace transmission
problem, however it appears to be valid for field scale problems, for
essentially the same reasons, as widely reported in the literature on
seismic inversion.

One (of many) alternative optimization formulation of this and similar
problems involves introducing an extension of the model-to-data
operator, and compensating for the physically spurious degrees of
freedom so introduced by adding a penalty term controlling them to the
objective.  The application of this idea presented here extends $\lF$
by dropping the support constraint on its second argument, that is,
using the expression \ref{eqn:mod} to define a map $F: M \times W
\rightarrow D$. The objective begins with an extended version of the
mean-square error,
\begin{equation}
  \label{eqn:mserr}
  e[m,w;d]=\frac{1}{2}\|F[m]w-d\|^2.
\end{equation}
We will choose a penalty to control the extra degrees of freedom, that
is, $w$ outside of $[-\lambda,\lambda]$. This  penalty is quadratic, constrains only $w\in W$ (as it is a soft
version of the condition $w \in \lW$), and is
$m-$independent. Therefore the penalty function takes the form
\begin{equation}
\label{eqn:pen}
p[w] = \frac{1}{2}\|Aw\|_Y^2 
\end{equation}
where $A \in {\cal B}(W,Y)$, $Y$ being another suitable Hilbert
space.

The penalty approach asks that a linear combination of $e,p$,
depending on a relative weight $\alpha$:
\begin{equation}
\Ja[m, w;d] = e[m,w;d] + \alpha^2 p[w] = \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)
\label{eqn:pen}
\end{equation}
be minimized over $(m,w)\in M \times W$.

The choice of penalty weight $\alpha$ has a profound influence on the
character of this optimization problem. For the moment, we will mandate
only that $\alpha \ge 0$. A complete description of an algorithm using
the penalty function \ref{eqn:pen} to solve the inverse problem
\ref{eqn:probstat0} must include a selection principle for $\alpha$. We
describe one such principle in Appendix A.

The main theoretical device used in the proofs of our main results on
extended inversion is the reduction of $\Ja$ to a function $\tJa$ of
$m$ alone, by minimization over $w$. This reduction is the foundation
of the {\em Variable Projection Method} (VPM)
\cite[]{GolubPereyra:03} for solution of partly linear (``separable'')
least squares problems. We use the VPM in our numerical illustrations
of the theory, as described in Discussion section and Appendix A.

\begin{theorem}
\label{thm:red}
Assume that the Hessian operator of
$\Ja[m,\cdot;d]$ is invertible:
\begin{equation}
\label{eqn:norminv}
F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
\end{equation}
(The superscript $T$ denotes the adjoint with respect to the inner products 
in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
Then the unique stationary point $\aw[m;d] \in W$ of
$\Ja[m,\cdot;d]$ is the solution of the {\em normal equation}:
\begin{equation}
  \label{eqn:norm}
  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td,
\end{equation}
Define the reduced or variable projection objective by
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
\end{equation}
Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
\end{theorem}
\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$ are differentiable, then
  ``local minimizer'' in the conclusion of the preceding theorem can
  be replaced by ``stationary point''. However, for the problem
  addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is not}
  differentiable without added smoothness constraints on $w$, whereas
  $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of penalty
  operator $A$, as will be shown
  below. This phenomenon is replicated for a wide variety of extended
  approaches for inverse problems in wave propagation. We will have
  more to say about it in the Discussion section.
